---
title: "Group Project"
author: "Keyan Cummins"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT
This study investigates the relationship between state-level outdoor air quality and adult asthma prevalence across the 50 U.S. states. Using simple linear regression and Pearson correlation, the analysis tested the hypothesis that states with a higher percentage of poor air quality days would exhibit higher asthma rates. The data revealed a weak, statistically significant negative correlation (r = -0.279, p = 0.0497) between poor air quality days and asthma prevalence. This finding indicates that outdoor air quality accounts for less than 8% of the variability in state asthma rates and that the overall prevalence is more likely driven by powerful confounding factors such as indoor air quality, public health practices, or regional climate. The results reject the initial hypothesis and suggest that a simple, direct link between state-level outdoor air quality and asthma prevalence is not supported by this dataset.

# BACKGROUND
Asthma is a lung condition that causes difficulty in breathing due to constricted airways in the lungs. There are many different kinds of asthma such as: exercise-induced asthma, occupational asthma, and allergy-induced asthma (Asthma). Those with occupational asthma can have attacks triggered by things such as gases, fumes, and dust, similar to that of air pollution (Asthma). Severity of the condition differs from person to person with some only experiencing minor symptoms while others can experience life-threatening challenges from the disease (Asthma). Asthma is a very common disease in the U.S., with about 28 million people in the U.S. suffering from asthma (Asthma Facts).

Poor air quality can impact those who suffer from asthma by causing their condition to worsen and trigger attacks (Tiotiu). It has been found in epidemiological studies that greater levels of air pollution has been associated with childhood asthma and that following an improvement of air quality, childhood asthma incidence had lowered (Jama). The state with the worst air quality in the U.S. is California due to its high air pollution.

It is believed that air quality can affect the prevalence of asthma per state. We hypothesize that states that have a higher amount of poor air quality days will have a higher prevalence of asthma. We predict that California will have the most poor air quality days and, because of that, they will also have the highest rates of asthma compared to other states. We took data from the United States Environmental Protection Agency on asthma prevelance and air quality index by state. To ensure the validity of our statistical approach, we utilized histograms and a Shapiro-Wilk test to determine if the data was normally distributed before proceeding with correlation and regression analysis.

```{r data}
# Load the data
air_data <- data.frame(
  State = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"),
  Poor_Air = c(0.27, 0.12, 0.33, 0.3, 0.46, 0.25, 0.22, 0.31, 0.24, 0.3, 0.03, 0.28, 0.34, 0.25, 0.27, 0.29, 0.23, 0.29, 0.08, 0.19, 0.23, 0.24, 0.14, 0.31, 0.22, 0.19, 0.1, 0.25, 0.1, 0.27, 0.22, 0.16, 0.2, 0.17, 0.25, 0.34, 0.29, 0.31, 0.16, 0.25, 0.15, 0.23, 0.33, 0.29, 0.13, 0.12, 0.24, 0.16, 0.2, 0.15),
  Asthma = c(9.6, 10.7, 9.7, 10.5, 8.7, 10.8, 12.4, 9.9, 9.3, 9.6, 9.1, 11.1, 8.7, 11, 9.7, 10.7, 10.8, 10, 13.1, 10.4, 11.3, 11.9, 9.9, 9.4, 10.4, 11.7, 8.1, 10.1, 13.1, 8.9, 10.4, 10.3, 9.2, 10.4, 11.4, 12.3, 11.5, 10.1, 13.3, 9, 8.3, 11.7, 7.9, 11, 12.9, 9.9, 10.9, 12.9, 10.9, 10.5)
)
```

```{r asthma-ranked-bars, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=12, fig.cap="Adults with Current Asthma vs. Poor-Air Days by State (ranked by Asthma %)"}
# Use the existing air_data from our environment
# Convert to appropriate format and calculate percentages
df <- data.frame(
  State = air_data$State,
  asthma_pct = air_data$Asthma,
  poor_air_pct = air_data$Poor_Air * 100  # Convert proportion to percentage
)

# Sort by asthma percentage (descending)
df <- df[order(-df$asthma_pct), ]

# Set up plotting parameters
par(mar = c(5, 10, 4, 6))  # Adjust margins for state names
y_pos <- 1:nrow(df)  # Y positions for bars

# Create empty plot
plot(0, 0, type = "n", 
     xlim = c(0, 50), 
     ylim = c(0.5, nrow(df) + 0.5),
     xlab = "Percent", 
     ylab = "",
     main = "Adults with Current Asthma vs. Poor-Air Days by State\n(Ranked by Asthma %)",
     yaxt = "n")  # No default y-axis

# Add asthma bars (blue)
barplot_height <- df$asthma_pct
rect(xleft = 0, ybottom = y_pos - 0.3, 
     xright = barplot_height, ytop = y_pos + 0.3,
     col = "lightblue", border = NA)

# Add air quality bars (red) - offset slightly
barplot_height_air <- df$poor_air_pct
rect(xleft = 0, ybottom = y_pos - 0.3 + 0.6, 
     xright = barplot_height_air, ytop = y_pos + 0.3 + 0.6,
     col = "lightcoral", border = NA)

# Add state names
axis(2, at = y_pos, labels = df$State, las = 1, cex.axis = 0.6)

# Add legend
legend("bottomright", 
       legend = c("Adults with Current Asthma (%)", "Poor Air Quality Days (%)"),
       fill = c("lightblue", "lightcoral"),
       cex = 0.8,
       bty = "n")
# This code was edited by prompts I gave to ChatGTP
```

```{r}
# Set up side-by-side plots
par(mfrow = c(1, 2), mar = c(5, 4, 4, 2))

# Histogram with density curve for Asthma
hist(df$asthma_pct, 
     breaks = 15,
     main = "Distribution of Asthma Prevalence",
     xlab = "Adults with Current Asthma (%)",
     ylab = "Density",
     col = "lightblue",
     border = "darkblue",
     freq = FALSE,
     xlim = c(0, max(df$asthma_pct) + 2))

# Add density curve
lines(density(df$asthma_pct), col = "darkblue", lwd = 2)


# Histogram with density curve for Poor Air Quality
hist(df$poor_air_pct, 
     breaks = 15,
     main = "Distribution of Poor Air Quality Days",
     xlab = "Poor Air Quality Days (%)",
     ylab = "Density",
     col = "lightcoral",
     border = "darkred",
     freq = FALSE,
     xlim = c(0, max(df$poor_air_pct) + 5))

# Add density curve
lines(density(df$poor_air_pct), col = "darkred", lwd = 2)


# Reset plotting parameters
par(mfrow = c(1, 1))

# This code was edited by prompts I gave to ChatGTP
```

```{r}
# Shapiro-Wilk tests for normality
shapiro_poor_air <- shapiro.test(air_data$Poor_Air)
shapiro_asthma <- shapiro.test(air_data$Asthma)

# Create results table
normality_tests <- data.frame(
  Variable = c("Poor Air Quality Days", "Asthma Rate"),
  Statistic_W = c(round(shapiro_poor_air$statistic, 4), 
                  round(shapiro_asthma$statistic, 4)),
  P_Value = c(ifelse(shapiro_poor_air$p.value < 0.001, "< 0.001", 
                     round(shapiro_poor_air$p.value, 4)),
              ifelse(shapiro_asthma$p.value < 0.001, "< 0.001", 
                     round(shapiro_asthma$p.value, 4))),
  Normal_Distribution = c(ifelse(shapiro_poor_air$p.value > 0.05, "Yes", "No"),
                          ifelse(shapiro_asthma$p.value > 0.05, "Yes", "No"))
)

# Display results
knitr::kable(normality_tests, caption = "Shapiro-Wilk Normality Tests")

# This code was edited by prompts I gave to ChatGTP
```


# STUDY QUESTION and HYPOTHESIS

## Questions  
Does air quality effect the prevalence of asthma per state?

## Hypothesis 
The states with the highest amounts of poor air quality days will also have higher prevalence of asthma.
  
## Prediction
California will have the highest prevelance of asthma because of its' high amounts of poor air quality days. 

# METHODS/RESULTS

## 1st Analysis (Linear Regression)
The first analysis utilized Simple Linear Regression to determine the presence, strength, and direction of a linear relationship between the predictor variable (Proportion of Poor Air Quality Days) and the response variable (Asthma Rate in percent).

```{r}
## 1st Analysis: Linear Regression - Air Quality vs Asthma Rates

# Convert Poor_Air from proportion to percentage
air_data$Poor_Air_Percent <- air_data$Poor_Air * 100

# Scatter plot with regression line
plot(air_data$Poor_Air_Percent, air_data$Asthma,
     xlab = "Percentage of Poor Air Quality Days",
     ylab = "Asthma Rate (%)",
     main = "Air Quality vs Asthma Rates by State",
     pch = 16, col = "blue", cex = 1.2)

# Add regression line
model <- lm(Asthma ~ Poor_Air_Percent, data = air_data)
abline(model, col = "red", lwd = 2)

# Calculate statistics
correlation <- cor(air_data$Poor_Air_Percent, air_data$Asthma)
r_squared <- summary(model)$r.squared
p_value <- summary(model)$coefficients[2,4]

# Create comprehensive results table
regression_stats <- data.frame(
  Statistic = c("Correlation Coefficient", "R-squared", "Adjusted R-squared", 
                "Overall Model p-value", "Regression p-value", "Observations"),
  Value = c(round(correlation, 3),
            round(r_squared, 3),
            round(summary(model)$adj.r.squared, 3),
            round(pf(summary(model)$fstatistic[1], summary(model)$fstatistic[2], 
                     summary(model)$fstatistic[3], lower.tail = FALSE), 4),
            round(p_value, 4),
            nrow(air_data))
)


# Create coefficients table
coefficients_table <- data.frame(
  Term = c("Intercept", "Poor Air Quality (%)"),
  Estimate = c(round(coef(model)[1], 3), round(coef(model)[2], 3)),
  Std_Error = c(round(summary(model)$coefficients[1,2], 3), 
                round(summary(model)$coefficients[2,2], 3)),
  t_value = c(round(summary(model)$coefficients[1,3], 3), 
              round(summary(model)$coefficients[2,3], 3)),
  p_value = c(round(summary(model)$coefficients[1,4], 4), 
              round(summary(model)$coefficients[2,4], 4))
)


# Add clean statistics to plot
legend("topright", 
       legend = c(paste("r =", round(correlation, 3)),
                  paste("R² =", round(r_squared, 3)),
                  paste("p =", ifelse(p_value < 0.001, "< 0.001", round(p_value, 3)))),
       bty = "n", cex = 0.9)

# This code was edited by prompts I gave to ChatGTP
```


## 2nd Analysis (Pearson Correlation)
The second analysis utilized Pearson Correlation to formally test the null hypothesis that there is no linear relationship between the two variables.

```{r}

## Pearson Correlation Analysis

# Calculate Pearson correlation
cor_test <- cor.test(air_data$Poor_Air, air_data$Asthma, method = "pearson")

# Create correlation results table
correlation_table <- data.frame(
  Statistic = c("Pearson Correlation Coefficient", 
                "95% Confidence Interval Lower",
                "95% Confidence Interval Upper",
                "t-statistic",
                "Degrees of Freedom",
                "p-value",
                "Sample Size"),
  Value = c(round(cor_test$estimate, 3),
            round(cor_test$conf.int[1], 3),
            round(cor_test$conf.int[2], 3),
            round(cor_test$statistic, 3),
            cor_test$parameter,
            ifelse(cor_test$p.value < 0.001, "< 0.001", round(cor_test$p.value, 4)),
            nrow(air_data))
)

# Display correlation results
knitr::kable(correlation_table, caption = "Pearson Correlation Analysis")

# This code was edited by prompts I gave to ChatGTP
```


# DISCUSSION

## Interpretation of 1st analysis 
The Simple Linear Regression analysis established a model to predict the Asthma Rate based on the Percentage of Poor Air Quality Days. The analysis revealed a Correlation Coefficient of r = -0.279, indicating a weak negative linear relationship. This suggests that as the percentage of poor air quality days increases, the asthma rate tends to slightly decrease. The Coefficient of Determination R^2 was 0.078, meaning that approximately 7.8% of the variability in asthma rates across the 50 states can be explained by poor air quality days. Importantly, the Regression p-value was 0.0497, which is just below the significance level of alpha = 0.05. Therefore, we reject the null hypothesis, concluding that there is a statistically significant linear relationship between these variables. However, the direction of this relationship (negative) is the opposite of what was expected.

## Interpretation of 2nd analysis 
The Pearson Correlation analysis reinforced the findings from the regression. The Pearson correlation coefficient r = -0.279 confirmed the weak negative association. The calculated p-value of 0.0497 indicates that the result is statistically significant at the 95% confidence level. This means that the likelihood of observing this relationship by random chance (if there were truly no relationship) is less than 5%. Consequently, we have statistical evidence to suggest a non-zero correlation, but the result directly contradicts our prediction. The data indicate that states with higher frequencies of poor air quality days (like California) actually report lower percentages of adults with current asthma compared to states with better air quality (like Maine or New Hampshire).

# CONCLUSION
The study found a statistically significant but weak negative linear relationship r = -0.279, p = 0.0497 between the percentage of poor air quality days and adult asthma prevalence across the 50 US states. While the results are statistically significant, they reject the initial hypothesis and prediction. We predicted a positive relationship where poor air quality leads to higher asthma rates, but the data shows the opposite. These findings suggests that state-level asthma prevalence is likely caused by other powerful factors that may be related to the asthma rate, such as socioeconomic status, access to healthcare, smoking rates, or potentially the "healthy worker effect" where individuals with respiratory issues might move away from areas with known poor air quality. Future research should look at some of these confounding variables to see the trends they have.

# REFERENCES     

1. “Asthma.” Mayo Clinic, Mayo Foundation for Medical Education and Research, 8 Mar. 2025, www.mayoclinic.org/diseases-conditions/asthma/symptoms-causes/syc-20369653.

2. “Asthma Facts.” Asthma & Allergy Foundation of America, 23 Apr. 2025, aafa.org/asthma/asthma-facts/.

3. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed r Sys.Date().

4. Google. (2025). Gemini (version Oct 2025). Tool used for quick fixes, editing grammar and flow of text, and checking all rubric requirements were met.

5. “Jama Network | Home of Jama and the Specialty Journals of the American Medical Association.” Jama Network, jamanetwork.com/. Accessed 12 Nov. 2025.

6. Tiotiu AI, Novakova P, Nedeva D, Chong-Neto HJ, Novakova S, Steiropoulos P, Kowal K. Impact of Air Pollution on Asthma Outcomes. Int J Environ Res Public Health. 2020 Aug 27;17(17):6212. doi: 10.3390/ijerph17176212. PMID: 32867076; PMCID: PMC7503605.

7. U.S. Environmental Protection Agency (EPA). Asthma and Outdoor Air Pollution [flyer]. EPA-452-F-04-002. Used for background information on how outdoor air pollution affects asthma and suggested protective behaviors. Accessed r Sys.Date(). 
