---
title: "Group Project"
author: "Keyan Cummins"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT


# BACKGROUND
Asthma is a chronic respiratory condition characterized by inflammation and constriction of the airways, leading to breathing difficulties. Its severity varies significantly among individuals, ranging from a minor disease to a potentially life-threatening disease. For those with asthma, poor air quality can act as a trigger, exacerbating symptoms and increasing the frequency of attacks (Tiotiu). This relationship suggests that air quality may influence the geographic prevalence of asthma. We hypothesize that states with a greater number of poor air quality days will demonstrate a higher prevalence of asthma. We predict that California, with the most poor air quality days, will also have the highest asthma rates in the nation.

```{r data}
# Load the data
air_data <- data.frame(
  State = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"),
  Poor_Air = c(0.27, 0.12, 0.33, 0.3, 0.46, 0.25, 0.22, 0.31, 0.24, 0.3, 0.03, 0.28, 0.34, 0.25, 0.27, 0.29, 0.23, 0.29, 0.08, 0.19, 0.23, 0.24, 0.14, 0.31, 0.22, 0.19, 0.1, 0.25, 0.1, 0.27, 0.22, 0.16, 0.2, 0.17, 0.25, 0.34, 0.29, 0.31, 0.16, 0.25, 0.15, 0.23, 0.33, 0.29, 0.13, 0.12, 0.24, 0.16, 0.2, 0.15),
  Asthma = c(9.6, 10.7, 9.7, 10.5, 8.7, 10.8, 12.4, 9.9, 9.3, 9.6, 9.1, 11.1, 8.7, 11, 9.7, 10.7, 10.8, 10, 13.1, 10.4, 11.3, 11.9, 9.9, 9.4, 10.4, 11.7, 8.1, 10.1, 13.1, 8.9, 10.4, 10.3, 9.2, 10.4, 11.4, 12.3, 11.5, 10.1, 13.3, 9, 8.3, 11.7, 7.9, 11, 12.9, 9.9, 10.9, 12.9, 10.9, 10.5)
)
```

```{r asthma-ranked-bars, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=12, fig.cap="Adults with Current Asthma vs. Poor-Air Days by State (ranked by Asthma %)"}
# Use the existing air_data from our environment
# Convert to appropriate format and calculate percentages
df <- data.frame(
  State = air_data$State,
  asthma_pct = air_data$Asthma,
  poor_air_pct = air_data$Poor_Air * 100  # Convert proportion to percentage
)

# Sort by asthma percentage (descending)
df <- df[order(-df$asthma_pct), ]

# Set up plotting parameters
par(mar = c(5, 10, 4, 6))  # Adjust margins for state names
y_pos <- 1:nrow(df)  # Y positions for bars

# Create empty plot
plot(0, 0, type = "n", 
     xlim = c(0, 50), 
     ylim = c(0.5, nrow(df) + 0.5),
     xlab = "Percent", 
     ylab = "",
     main = "Adults with Current Asthma vs. Poor-Air Days by State\n(Ranked by Asthma %)",
     yaxt = "n")  # No default y-axis

# Add asthma bars (blue)
barplot_height <- df$asthma_pct
rect(xleft = 0, ybottom = y_pos - 0.3, 
     xright = barplot_height, ytop = y_pos + 0.3,
     col = "lightblue", border = NA)

# Add air quality bars (red) - offset slightly
barplot_height_air <- df$poor_air_pct
rect(xleft = 0, ybottom = y_pos - 0.3 + 0.6, 
     xright = barplot_height_air, ytop = y_pos + 0.3 + 0.6,
     col = "lightcoral", border = NA)

# Add state names
axis(2, at = y_pos, labels = df$State, las = 1, cex.axis = 0.6)

# Add legend
legend("bottomright", 
       legend = c("Adults with Current Asthma (%)", "Poor Air Quality Days (%)"),
       fill = c("lightblue", "lightcoral"),
       cex = 0.8,
       bty = "n")
```

```{r}
# Set up side-by-side plots
par(mfrow = c(1, 2), mar = c(5, 4, 4, 2))

# Histogram with density curve for Asthma
hist(df$asthma_pct, 
     breaks = 15,
     main = "Distribution of Asthma Prevalence",
     xlab = "Adults with Current Asthma (%)",
     ylab = "Density",
     col = "lightblue",
     border = "darkblue",
     freq = FALSE,
     xlim = c(0, max(df$asthma_pct) + 2))

# Add density curve
lines(density(df$asthma_pct), col = "darkblue", lwd = 2)


# Histogram with density curve for Poor Air Quality
hist(df$poor_air_pct, 
     breaks = 15,
     main = "Distribution of Poor Air Quality Days",
     xlab = "Poor Air Quality Days (%)",
     ylab = "Density",
     col = "lightcoral",
     border = "darkred",
     freq = FALSE,
     xlim = c(0, max(df$poor_air_pct) + 5))

# Add density curve
lines(density(df$poor_air_pct), col = "darkred", lwd = 2)


# Reset plotting parameters
par(mfrow = c(1, 1))
```

```{r}
# Shapiro-Wilk tests for normality
shapiro_poor_air <- shapiro.test(air_data$Poor_Air)
shapiro_asthma <- shapiro.test(air_data$Asthma)

# Create results table
normality_tests <- data.frame(
  Variable = c("Poor Air Quality Days", "Asthma Rate"),
  Statistic_W = c(round(shapiro_poor_air$statistic, 4), 
                  round(shapiro_asthma$statistic, 4)),
  P_Value = c(ifelse(shapiro_poor_air$p.value < 0.001, "< 0.001", 
                     round(shapiro_poor_air$p.value, 4)),
              ifelse(shapiro_asthma$p.value < 0.001, "< 0.001", 
                     round(shapiro_asthma$p.value, 4))),
  Normal_Distribution = c(ifelse(shapiro_poor_air$p.value > 0.05, "Yes", "No"),
                          ifelse(shapiro_asthma$p.value > 0.05, "Yes", "No"))
)

# Display results
knitr::kable(normality_tests, caption = "Shapiro-Wilk Normality Tests")
```


# STUDY QUESTION and HYPOTHESIS

## Questions  
Does air quality effect the prevalence of asthma per state?

## Hypothesis 
The states with the highest amounts of poor air quality days will also have higher prevalence of asthma.
  
## Prediction
California will have the highest prevelance of asthma because of its' high amounts of poor air quality days. 

# METHODS/RESULTS

## 1st Analysis 
The first analysis utilized Simple Linear Regression to determine the presence, strength, and direction of a linear relationship between the predictor variable (Proportion of Poor Air Quality Days) and the response variable (Asthma Rate in percent).

```{r}
## 1st Analysis: Linear Regression - Air Quality vs Asthma Rates

# Convert Poor_Air from proportion to percentage
air_data$Poor_Air_Percent <- air_data$Poor_Air * 100

# Scatter plot with regression line
plot(air_data$Poor_Air_Percent, air_data$Asthma,
     xlab = "Percentage of Poor Air Quality Days",
     ylab = "Asthma Rate (%)",
     main = "Air Quality vs Asthma Rates by State",
     pch = 16, col = "blue", cex = 1.2)

# Add regression line
model <- lm(Asthma ~ Poor_Air_Percent, data = air_data)
abline(model, col = "red", lwd = 2)

# Calculate statistics
correlation <- cor(air_data$Poor_Air_Percent, air_data$Asthma)
r_squared <- summary(model)$r.squared
p_value <- summary(model)$coefficients[2,4]

# Create comprehensive results table
regression_stats <- data.frame(
  Statistic = c("Correlation Coefficient", "R-squared", "Adjusted R-squared", 
                "Overall Model p-value", "Regression p-value", "Observations"),
  Value = c(round(correlation, 3),
            round(r_squared, 3),
            round(summary(model)$adj.r.squared, 3),
            round(pf(summary(model)$fstatistic[1], summary(model)$fstatistic[2], 
                     summary(model)$fstatistic[3], lower.tail = FALSE), 4),
            round(p_value, 4),
            nrow(air_data))
)


# Create coefficients table
coefficients_table <- data.frame(
  Term = c("Intercept", "Poor Air Quality (%)"),
  Estimate = c(round(coef(model)[1], 3), round(coef(model)[2], 3)),
  Std_Error = c(round(summary(model)$coefficients[1,2], 3), 
                round(summary(model)$coefficients[2,2], 3)),
  t_value = c(round(summary(model)$coefficients[1,3], 3), 
              round(summary(model)$coefficients[2,3], 3)),
  p_value = c(round(summary(model)$coefficients[1,4], 4), 
              round(summary(model)$coefficients[2,4], 4))
)


# Add clean statistics to plot
legend("topright", 
       legend = c(paste("r =", round(correlation, 3)),
                  paste("R² =", round(r_squared, 3)),
                  paste("p =", ifelse(p_value < 0.001, "< 0.001", round(p_value, 3)))),
       bty = "n", cex = 0.9)
```




## 2nd Analysis 

```{r}

## Pearson Correlation Analysis

# Calculate Pearson correlation
cor_test <- cor.test(air_data$Poor_Air, air_data$Asthma, method = "pearson")

# Create correlation results table
correlation_table <- data.frame(
  Statistic = c("Pearson Correlation Coefficient", 
                "95% Confidence Interval Lower",
                "95% Confidence Interval Upper",
                "t-statistic",
                "Degrees of Freedom",
                "p-value",
                "Sample Size"),
  Value = c(round(cor_test$estimate, 3),
            round(cor_test$conf.int[1], 3),
            round(cor_test$conf.int[2], 3),
            round(cor_test$statistic, 3),
            cor_test$parameter,
            ifelse(cor_test$p.value < 0.001, "< 0.001", round(cor_test$p.value, 4)),
            nrow(air_data))
)

# Display correlation results
knitr::kable(correlation_table, caption = "Pearson Correlation Analysis")
```




# DISCUSSION

## Interpretation of 1st analysis 


## Interpretation of 2nd analysis 


# CONCLUSION
The analyses performed in this study failed to support the hypothesis that states with a higher number of poor air quality days have a higher prevalence of asthma. Instead, the data presented a statistically significant, yet weak, negative correlation. Specifically, the Low Pollution group was found to have a significantly higher mean asthma rate than the High Pollution group.

The prediction that California (which has one of the highest proportions of poor air days (at 46%) would have the highest asthma prevalence was empirically incorrect; California has one of the lowest asthma rates (8.7%). Conversely, states like Rhode Island, New Hampshire, and Vermont, which generally have lower poor air days, report some of the highest asthma rates (up to 13.3%).

This evidence strongly suggests that state-level outdoor air quality is not the dominant factor determining asthma prevalence. At this broad geographic scale, asthma rates are likely more strongly influenced by powerful confounding variables, such as:

1. Indoor Environment: Factors like poor ventilation, mold, dust, and smoking within the home.

2. Climate and Geography: Humidity, temperature, and the intensity of seasonal allergens.

3. Public Health Practices and Reporting: Regional differences in diagnosis, health education, and public health campaigns.

Future studies should use individual-level data or employ multivariate regression techniques that can isolate the effect of air quality while controlling for confounding demographic and climate variables.

# REFERENCES     

1. Tiotiu AI, Novakova P, Nedeva D, Chong-Neto HJ, Novakova S, Steiropoulos P, Kowal K. Impact of Air Pollution on Asthma Outcomes. Int J Environ Res Public Health. 2020 Aug 27;17(17):6212. doi: 10.3390/ijerph17176212. PMID: 32867076; PMCID: PMC7503605.

3. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.

4. Google. (2025). Gemini (version Oct 2025). Tool used for quick fixes, editing grammar and flow of text, and checking all rubric requirements were met.