---
title: "Group Project"
author: "Keyan Cummins"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

# BACKGROUND


```{r}
# Load the data
air_data <- data.frame(
  State = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"),
  Poor_Air = c(0.27, 0.12, 0.33, 0.3, 0.46, 0.25, 0.22, 0.31, 0.24, 0.3, 0.03, 0.28, 0.34, 0.25, 0.27, 0.29, 0.23, 0.29, 0.08, 0.19, 0.23, 0.24, 0.14, 0.31, 0.22, 0.19, 0.1, 0.25, 0.1, 0.27, 0.22, 0.16, 0.2, 0.17, 0.25, 0.34, 0.29, 0.31, 0.16, 0.25, 0.15, 0.23, 0.33, 0.29, 0.13, 0.12, 0.24, 0.16, 0.2, 0.15),
  Asthma = c(9.6, 10.7, 9.7, 10.5, 8.7, 10.8, 12.4, 9.9, 9.3, 9.6, 9.1, 11.1, 8.7, 11, 9.7, 10.7, 10.8, 10, 13.1, 10.4, 11.3, 11.9, 9.9, 9.4, 10.4, 11.7, 8.1, 10.1, 13.1, 8.9, 10.4, 10.3, 9.2, 10.4, 11.4, 12.3, 11.5, 10.1, 13.3, 9, 8.3, 11.7, 7.9, 11, 12.9, 9.9, 10.9, 12.9, 10.9, 10.5)
)
```

```{r}
# Scatter plot with regression line and statistics
plot(air_data$Poor_Air, air_data$Asthma,
     xlab = "Proportion of Poor Air Quality Days",
     ylab = "Asthma Rate (%)",
     main = "Air Quality vs Asthma Rates by State",
     pch = 16, col = "blue", cex = 1.2)

# Add regression line
model <- lm(Asthma ~ Poor_Air, data = air_data)
abline(model, col = "red", lwd = 2)

# Calculate statistics
correlation <- cor(air_data$Poor_Air, air_data$Asthma)
r_squared <- summary(model)$r.squared
p_value <- summary(model)$coefficients[2,4]

# Add statistics to plot
legend("topright", 
       legend = c(paste("Correlation:", round(correlation, 3)),
                  paste("R-squared:", round(r_squared, 3)),
                  paste("p-value:", round(p_value, 4))),
       bty = "n")

# Print statistics to console
cat("=== REGRESSION STATISTICS ===\n")
cat("Correlation:", round(correlation, 3), "\n")
cat("R-squared:", round(r_squared, 3), "\n")
cat("p-value:", round(p_value, 4), "\n")
cat("Regression equation: Asthma =", 
    round(coef(model)[1], 2), "+", 
    round(coef(model)[2], 2), "* Poor_Air\n")
```


# STUDY QUESTION and HYPOTHESIS

## Questions  
Does air quality effect the prevalence of asthma per state?

## Hypothesis 
The states with the highest amounts of poor air quality days will also have higher prevalence of asthma.
  
## Prediction
California will have the highest prevelance of asthma because of its' high amounts of poor air quality days. 

# METHODS


## 1st Analysis 

```{r first-analysis, warning=FALSE}

```
## 2nd Analysis 

```{r}

```

# DISCUSSION

## Interpretation of 1st analysis 

## Interpretation of 2nd analysis 


# CONCLUSION


# REFERENCES     


3. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.

4. Google. (2025). Gemini (version Oct 2025). Tool used for quick fixes, editing grammar and flow of text, and checking all rubric requirements were met.