---
title: "Warm-up mini-Report: Mosquito Blood Hosts in \n Salt Lake City, Utah"
author: "Keyan Cummins"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

This research focused on analyzing Salt Lake County, Utah, to determine the host feeding habits of mosquitoes in order to find possible amplifying hosts for the West Nile virus (WNV) and to evaluate the possible high-risk areas for transmission. The blood meal data captured from mosquitoes within the given collection sites were assessed for risk WNV+ and WNV- locations, and host availability was compared for the differing sites. The barplots showed that the blood meals from the House Finch were most numerous overall and were disproportionately present in WNV+ sites. A generalized linear model showed that the number of House Finch blood meals was a strong, positive, and significant predictor of WNV rate in a given location. These results indicate that the House Finch is likely one of the primary WNV amplifying hosts in Salt Lake City. To control the spread of WNV, areas with high populations of House Finch should be the areas of focus. 

# BACKGROUND

West Nile virus (WNV) is a mosquito-borne disease maintained in a transmission cycle between mosquitoes and birds, which act as hosts for the virus. Since being introduced to North America in 1999, WNV has become a major public health concern, especially in the areas around Salt Lake City. WNV is commonly found in mosquito populations in this area [2]. Understanding which bird species mosquitoes feed on is essential for predicting where and how the virus spreads.

Studying mosquito feeding is crucial for predicting WNV spread. The main focus is identifying the main carriers, or amplifying hosts. These amplifying hosts are birds that maintain high concentrations of virus in their blood (high viremia) long enough to infect mosquitoes that feed on them. Research shows that species like the House Finch are highly competent hosts because they sustain high viremia, making them important drivers of WNV transmission [1].

Therefore, to understand and control local WNV transmission, it is essential to determine precisely which vertebrate hosts mosquitoes are feeding on. Researchers use molecular surveillance techniques to analyze mosquito blood meals. This typically involves DNA extraction to isolate host genetic material, followed by PCR (Polymerase Chain Reaction) to specifically amplify a short, identifiable segment of that host's DNA. Finally, the amplified DNA is analyzed through sequencing or other methods and compared to databases to identify the precise species, allowing for a targeted assessment of the risks posed by different species in the local environment.


```{r viremia, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# Choose some colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

# STUDY QUESTION and HYPOTHESIS

## Questions  
Which host species is the most common amplifying host for West Nile Virus in Salt Lake City, Utah?

## Hypothesis 
The House Finch will be acting as the most common amplifying host of WNV in Salt Lake City, Utah.
  
## Prediction
Locations where mosquitoes have taken blood meals from House Finches will also exhibit higher rates of WNV-positive mosquito pools compared to locations with less House Finch blood meals.

# METHODS
To figure out which birds are fueling the WNV cycle in the Salt Lake City area, we used molecular techniques. We extracted DNA from the mosquito's blood meal from multiple locations and then used Polymerase Chain Reaction (PCR) to amplify a target gene. After PCR, we sequenced the amplified DNA and referenced it to a database to identify the host species. Using the data from the identified hosts and the WNV positivity rates in the local mosquito pools helps us to understand more clearly the transmission of WNV in these areas.

The objective was to identify host species associated with increased West Nile virus (WNV) transmission risk in Salt Lake City. Mosquito blood meal host data from various sites were analyzed, with each site classified as either WNV-negative or WNV-positive.

## 1st Analysis (Barplots)
We counted the total number of blood meals taken from each host species. We added the counts separately for WNV-negative and WNV-positive locations. We visualized this comparison using horizontal bar plots to show host feeding patterns directly. The bar plots helped us see if any specific host was targeted more by mosquitoes in WNV-positive areas.

```{r first-analysis, warning=FALSE}
## import counts_matrix: data.frame with column 'loc_positives' (0/1) and host columns 'host_*'
counts_matrix <- read.csv("./bloodmeal_plusWNV_for_BIOL3070.csv")

## 1) Identify host columns
host_cols <- grep("^host_", names(counts_matrix), value = TRUE)

if (length(host_cols) == 0) {
  stop("No columns matching '^host_' were found in counts_matrix.")
}

## 2) Ensure loc_positives is present and has both levels 0 and 1 where possible
counts_matrix$loc_positives <- factor(counts_matrix$loc_positives, levels = c(0, 1))

## 3) Aggregate host counts by loc_positives
agg <- stats::aggregate(
  counts_matrix[, host_cols, drop = FALSE],
  by = list(loc_positives = counts_matrix$loc_positives),
  FUN = function(x) sum(as.numeric(x), na.rm = TRUE)
)

## make sure both rows exist; if one is missing, add a zero row
need_levels <- setdiff(levels(counts_matrix$loc_positives), as.character(agg$loc_positives))
if (length(need_levels)) {
  zero_row <- as.list(rep(0, length(host_cols)))
  names(zero_row) <- host_cols
  for (lv in need_levels) {
    agg <- rbind(agg, c(lv, zero_row))
  }
  ## restore proper type
  agg$loc_positives <- factor(agg$loc_positives, levels = c("0","1"))
  ## coerce numeric host cols (they may have become character after rbind)
  for (hc in host_cols) agg[[hc]] <- as.numeric(agg[[hc]])
  agg <- agg[order(agg$loc_positives), , drop = FALSE]
}

## 4) Decide species order (overall abundance, descending)
overall <- colSums(agg[, host_cols, drop = FALSE], na.rm = TRUE)
host_order <- names(sort(overall, decreasing = TRUE))
species_labels <- rev(sub("^host_", "", host_order))  # nicer labels

## 5) Build count vectors for each panel in the SAME order
counts0 <- rev(as.numeric(agg[agg$loc_positives == 0, host_order, drop = TRUE]))
counts1 <- rev(as.numeric(agg[agg$loc_positives == 1, host_order, drop = TRUE]))

## 6) Colors: reuse your existing 'cols' if it exists and is long enough; otherwise generate
if (exists("cols") && length(cols) >= length(host_order)) {
  species_colors <- setNames(cols[seq_along(host_order)], species_labels)
} else {
  species_colors <- setNames(rainbow(length(host_order) + 10)[seq_along(host_order)], species_labels)
}

## 7) Shared x-limit for comparability
xmax <- max(c(counts0, counts1), na.rm = TRUE)
xmax <- if (is.finite(xmax)) xmax else 1
xlim_use <- c(0, xmax * 1.08)

## 8) Plot: two horizontal barplots with identical order and colors
op <- par(mfrow = c(1, 2),
          mar = c(4, 12, 3, 2),  # big left margin for species names
          xaxs = "i")           # a bit tighter axis padding

## Panel A: No WNV detected (loc_positives = 0)
barplot(height = counts0,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (-)",
        xlim = xlim_use)

## Panel B: WNV detected (loc_positives = 1)
barplot(height = counts1,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (+)",
        xlim = xlim_use)

par(op)

## Keep the colors mapping for reuse elsewhere
host_species_colors <- species_colors
```
## 2nd Analysis (Generalized Linear Model)
To formally test the relationship shown in the barplots, we performed a Generalized Linear Model (GLM) with the number of House Finch blood meals as the independent variable. We ran two separate models. The first model used the binary presence of WNV (loc_positives) as the response variable with a binomial family. The second model used the WNV positivity rate (loc_rate) as the response variable with a Gaussian family. This statistical test gave us a formal p-value to assess the significance and strength of the House Finch as a predictor of WNV risk.

```{r}
# second-analysis-or-plot, glm with house finch alone against binary +/_
glm1 <- glm(loc_positives ~ host_House_finch,
            data = counts_matrix,
            family = binomial)
summary(glm1)

#glm with house-finch alone against positivity rate
glm2 <- glm(loc_rate ~ host_House_finch,
            data = counts_matrix)
summary(glm2)
```

# DISCUSSION

## Interpretation of 1st analysis (Barplots)
The first analysis, shown as a pair of bar charts, summarized the total blood meals by host species at WNV-negative and WNV-positive collection sites. The results showed that House Finch and American Robin were the most common blood meal hosts overall. In the WNV-positive locations (WNV+), the House Finch stood out as the main host, with blood meal counts much higher than any other species and higher than those seen in the WNV-negative locations. This visualization suggests that the House Finch is the main host connecting mosquito feeding activity to WNV transmission in the sampled areas.

A limitation of these bar plots is that they only show raw feeding counts, not host availability. The high number of House Finch blood meals in WNV+ sites might just be because there are more House Finches living there. Because we didn't measure the population of each bird species, the bar plots alone can't prove that mosquitoes are preferentially targeting House Finches.

## Interpretation of 2nd analysis (Generalized Linear Model)
The second analysis, a generalized linear model, confirmed the pattern seen in the barplots with statistics. The model that tested whether WNV was present showed a significant positive result for House Finch (p=0.0287). This indicates that locations with more House Finch blood meals were much more likely to have WNV-positive mosquito pools. Also, the model that examined the WNV positivity rate found a highly significant positive relationship (p=4.54×10^−5). This means that more House Finch blood meals lead to a statistically significant rise in the WNV infection rate at a location. Both models support the idea that House Finches play a key role in spreading WNV.

However, a limitation of this study is its correlational nature. While the strong statistical relationship suggests that the House Finch is a key host, the analysis does not prove direct causation. Unmeasured confounding factors, such as local environmental differences in habitat quality or hosts available to feed on, could influence both the mosquito feeding preference and the rate of WNV amplification at the same time.

# CONCLUSION
The conclusion we found for the research question is that the House Finch is likely the most important host for West Nile virus in Salt Lake City, Utah. The evidence strongly supports the hypothesis, showing that the number of House Finch blood meals is a significant positive predictor of WNV presence and positivity rate at a mosquito collection site. This finding aligns with biological data indicating that House Finches maintain a high viremia [1], confirming their role as a key host needed for the virus to spread effectively through the mosquito population in the Salt Lake City area. Public health measures should use this finding to prioritize mosquito control efforts in areas with large House Finch populations.

# REFERENCES     
1. Komar N, Langevin S, Hinten S, Nemeth N, Edwards E, Hettler D, Davis B, Bowen R, Bunning M. Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerg Infect Dis. 2003 Mar;9(3):311-22. https://doi.org/10.3201/eid0903.020628

2. Salt Lake County Health Department. (n.d.). West Nile virus detected in multiple mosquito samples in Salt Lake County. Retrieved from https://www.saltlakecounty.gov/health/news/west-nile-virus-detected-in-multiple-mosquito-samples-in-salt-lake-county/ 

3. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.

4. Google. (2025). Gemini (version Oct 2025). Tool used for quick fixes, editing grammar and flow of text, and checking all rubric requirements were met.